Current and future constraints on primordial magnetic fields

### 2016-10-28 pmf.tex
a

We present new limits on the amplitude of potential primordial magnetic fields (PMFs) using temperature and polarization measurements of the cosmic microwave background (CMB)  from \planck{}, \bicepkeck{}, \pb, and \sptpol. 
We reduce twofold the upper limit on the CMB anisotropy power due to a PMF, from $\apmf < 0.68$ from Planck alone to $\apmf < 0.33$ for the combined dataset at 95\% CL. 
%Most of the improvement is due to the addition of the \bicepkeck{} data; without these bandpowers, the combined limit weakens to $\apmf < 0.\tbd{XX}$. 
We also forecast the expected future limits from stage III CMB experiments (like \sptnew{},  \advactpol, or the \simons) and the proposed CMB stage IV experiment. 
Future CMB experiments should dramatically reduce the current uncertainties, by two orders of magnitude for stage III experiments and three orders of magnitude for stage IV. 
The constraints from a stage IV experiment have the potential to rule out much of the parameter space for PMFs.

@@

We find the addition of ground-based CMB polarization data to \planck{} data substantially improves the upper limits on the allowed power due to a PMF, leading to a factor of two improvement. 
With Planck alone, the 95\% confidence upper limit is $\apmf <  0.68$ when $\alens$ is allowed to vary; with the ground-based experiments, this limit drops to $\apmf <  0.33$. 
We test which datasets are important by adding single datasets to the \planck{} set, and find that the majority of this improvement is due to the \bicepkeck{} ($\apmf <  0.36$) and \sptpol{} ($\apmf <  0.42$) B-mode power spectrum measurements. 
The other datasets do not have a large impact on the observed limits.

We propagate these limits on the observed PMF power into limits on the magnetic field strength, \bpmf. 
Due to the scaling with the fourth power, an upper limit with a flat prior on \apmf{} would lead to an apparent `detection' of \bpmf. 
We therefor importance sample the chains to generate a flat prior on \bpmf. 
We find a 95\% limit of $\bpmf < XX$\,nG for \planck{} alone, consistent with \citep{planck15-19} with $n$ held fixed as we've done. 
This is reduced to $\bpmf < XX$\,nG for the combined dataset.

\begin{table*}[h]
\begin{center}
\caption{\label{tab:param_all} All parameter constraints}
\tiny
\begin{tabular}{l || c c c c c c c | c}
Chain & $\Omega_b h^2$  & $\Omega_c h^2$  & $\theta$  & $\tau$  & logA  & $n_s$  & $A_{lens}$  & $A_{pmf}$ \\
lcdm\_alens\_all & $ 0.02240\pm  0.00026$ & $ 0.1179\pm  0.0020$ & $ 1.04106\pm  0.00047$ & $ 0.064\pm  0.016$ & $ 3.057\pm  0.030$ & $ 0.9696\pm  0.0062$ & $ 1.121 \pm  0.064$ & $< 0.33$ \\
lcdm\_alens\_nopmf\_all & $ 0.02241\pm  0.00024$ & $ 0.1179\pm  0.0020$ & $ 1.04106\pm  0.00047$ & $ 0.065\pm  0.016$ & $ 3.059\pm  0.029$ & $ 0.9694\pm  0.0062$ & $ 1.136 \pm  0.063$ & $< 0.00$ \\
lcdm\_alens\_nopmf\_planck & $ 0.02256\pm  0.00027$ & $ 0.1173\pm  0.0021$ & $ 1.04121\pm  0.00048$ & $ 0.069\pm  0.017$ & $ 3.067\pm  0.030$ & $ 0.9721\pm  0.0063$ & $ 1.175 \pm  0.067$ & $< 0.00$ \\
lcdm\_alens\_planckTT & $ 0.02252\pm  0.00031$ & $ 0.1174\pm  0.0027$ & $ 1.04122\pm  0.00053$ & $ 0.070\pm  0.050$ & $ 3.068\pm  0.098$ & $ 0.9713\pm  0.0080$ & $ 1.155 \pm  0.136$ & $< 1.34$ \\
lcdm\_alens\_planckTT\_TEB & $ 0.02264\pm  0.00029$ & $ 0.1161\pm  0.0025$ & $ 1.04141\pm  0.00053$ & $ 0.055\pm  0.021$ & $ 3.036\pm  0.042$ & $ 0.9755\pm  0.0072$ & $ 1.241 \pm  0.104$ & $< 0.77$ \\
lcdm\_alens\_planck & $ 0.02254\pm  0.00026$ & $ 0.1173\pm  0.0020$ & $ 1.04121\pm  0.00046$ & $ 0.068\pm  0.016$ & $ 3.065\pm  0.030$ & $ 0.9723\pm  0.0062$ & $ 1.171 \pm  0.070$ & $< 0.68$ \\
lcdm\_alens\_planck\_bicep & $ 0.02255\pm  0.00026$ & $ 0.1174\pm  0.0020$ & $ 1.04120\pm  0.00047$ & $ 0.069\pm  0.017$ & $ 3.067\pm  0.030$ & $ 0.9724\pm  0.0063$ & $ 1.177 \pm  0.068$ & $< 0.36$ \\
lcdm\_alens\_planck\_pb & $ 0.02255\pm  0.00027$ & $ 0.1173\pm  0.0021$ & $ 1.04119\pm  0.00046$ & $ 0.068\pm  0.017$ & $ 3.066\pm  0.031$ & $ 0.9725\pm  0.0064$ & $ 1.171 \pm  0.068$ & $< 0.63$ \\
lcdm\_alens\_planck\_sptpolB & $ 0.02254\pm  0.00026$ & $ 0.1173\pm  0.0020$ & $ 1.04121\pm  0.00047$ & $ 0.069\pm  0.017$ & $ 3.067\pm  0.030$ & $ 0.9724\pm  0.0062$ & $ 1.159 \pm  0.067$ & $< 0.42$ \\
lcdm\_alens\_planck\_sptpolE & $ 0.02239\pm  0.00025$ & $ 0.1178\pm  0.0020$ & $ 1.04107\pm  0.00048$ & $ 0.064\pm  0.016$ & $ 3.057\pm  0.030$ & $ 0.9698\pm  0.0061$ & $ 1.119 \pm  0.067$ & $< 0.73$ \\
lcdm\_planck & $ 0.02225\pm  0.00023$ & $ 0.1184\pm  0.0020$ & $ 1.04102\pm  0.00047$ & $ 0.066\pm  0.017$ & $ 3.062\pm  0.030$ & $ 0.9682\pm  0.0060$ & $ 0.000 \pm  0.000$ & $< 0.73$ \\
lcdm\_r\_alens\_all & $ 0.02240\pm  0.00025$ & $ 0.1178\pm  0.0020$ & $ 1.04107\pm  0.00046$ & $ 0.064\pm  0.016$ & $ 3.056\pm  0.029$ & $ 0.9702\pm  0.0062$ & $ 1.123 \pm  0.063$ & $< 0.27$ \\
\end{tabular}
\tablecomments{ 
Summary of results (will be trimmed for paper). 
} \normalsize
\end{center}
\end{table*}

\begin{table*}[h]
\begin{center}
\caption{\label{tab:param_all2} All parameter constraints2}
\tiny
\begin{tabular}{l || c c c c c c c | c}
Chain & $\Omega_b h^2$  & $\Omega_c h^2$  & $\theta$  & $\tau$  & logA  & $n_s$  & $A_{lens}$  & $A_{pmf}$ \\
lcdm\_alens\_all\_bk14 & $ 0.02240\pm  0.00025$ & $ 0.1179\pm  0.0020$ & $ 1.04106\pm  0.00045$ & $ 0.065\pm  0.017$ & $ 3.058\pm  0.030$ & $ 0.9696\pm  0.0060$ & $ 1.131 \pm  0.061$ & $< 0.27$ \\
lcdm\_alens\_all\_bk14\_tensor & $ 0.02243\pm  0.00025$ & $ 0.1178\pm  0.0020$ & $ 1.04110\pm  0.00047$ & $ 0.066\pm  0.017$ & $ 3.060\pm  0.030$ & $ 0.9704\pm  0.0062$ & $ 1.142 \pm  0.060$ & $< 0.29$ \\
lcdm\_alens\_all\_bk14\_vector & $ 0.02240\pm  0.00025$ & $ 0.1181\pm  0.0019$ & $ 1.04107\pm  0.00046$ & $ 0.064\pm  0.016$ & $ 3.056\pm  0.029$ & $ 0.9687\pm  0.0061$ & $ 1.134 \pm  0.061$ & $< 0.68$ \\
lcdm\_alens\_nopmf\_all\_bk14 & $ 0.02244\pm  0.00025$ & $ 0.1179\pm  0.0020$ & $ 1.04109\pm  0.00046$ & $ 0.066\pm  0.016$ & $ 3.061\pm  0.029$ & $ 0.9700\pm  0.0060$ & $ 1.145 \pm  0.061$ & $< 0.00$ \\
lcdm\_alens\_planck\_bk14 & $ 0.02256\pm  0.00027$ & $ 0.1174\pm  0.0020$ & $ 1.04120\pm  0.00048$ & $ 0.069\pm  0.017$ & $ 3.068\pm  0.030$ & $ 0.9723\pm  0.0062$ & $ 1.188 \pm  0.066$ & $< 0.28$ \\
lcdm\_alens\_planck\_bk14\_tensor & $ 0.02257\pm  0.00026$ & $ 0.1173\pm  0.0020$ & $ 1.04121\pm  0.00047$ & $ 0.069\pm  0.017$ & $ 3.067\pm  0.030$ & $ 0.9727\pm  0.0060$ & $ 1.186 \pm  0.065$ & $< 0.29$ \\
lcdm\_alens\_planck\_bk14\_vector & $ 0.02241\pm  0.00029$ & $ 0.1180\pm  0.0021$ & $ 1.04114\pm  0.00048$ & $ 0.065\pm  0.017$ & $ 3.058\pm  0.030$ & $ 0.9670\pm  0.0074$ & $ 1.149 \pm  0.075$ & $< 4.64$ \\
lcdm\_all\_bk14 & $ 0.02223\pm  0.00021$ & $ 0.1182\pm  0.0018$ & $ 1.04113\pm  0.00049$ & $ 0.074\pm  0.014$ & $ 3.074\pm  0.024$ & $ 0.9686\pm  0.0055$ & $ 0.000 \pm  0.000$ & $< 0.35$ \\
lcdm\_all\_bk14\_tensor & $ 0.02220\pm  0.00024$ & $ 0.1186\pm  0.0019$ & $ 1.04100\pm  0.00047$ & $ 0.067\pm  0.016$ & $ 3.063\pm  0.029$ & $ 0.9672\pm  0.0057$ & $ 0.000 \pm  0.000$ & $< 0.31$ \\
lcdm\_all\_bk14\_vector & $ 0.02215\pm  0.00022$ & $ 0.1188\pm  0.0020$ & $ 1.04088\pm  0.00046$ & $ 0.064\pm  0.016$ & $ 3.056\pm  0.028$ & $ 0.9653\pm  0.0055$ & $ 0.000 \pm  0.000$ & $< 0.83$ \\
lcdm\_r\_alens\_all\_bk14 & $ 0.02242\pm  0.00025$ & $ 0.1175\pm  0.0021$ & $ 1.04111\pm  0.00048$ & $ 0.068\pm  0.018$ & $ 3.063\pm  0.032$ & $ 0.9710\pm  0.0063$ & $ 1.133 \pm  0.062$ & $< 0.24$ \\
\end{tabular}
\tablecomments{ 
Summary of results (will be trimmed for paper). 
} \normalsize
\end{center}
\end{table*}

@@

\begin{table*}[h]
\begin{center}
\caption{\label{tab:experiments} Assumed survey parameters}
\small
\begin{tabular}{l || c c c c c }
Experiment & Sky coverage & Polarized Noise level  & 1/f knee & Beam FWHM \\
& &($\mu$K-arcmin)&&(arcmin.)\\
\hline
Stage III: & & & & \\

@@

As a sanity check, we look at the Fisher matrix forecasts for Planck alone as well in both of these cases. 
For both chains,  we look at the sigma in MCMC with \apmf{} allowed to go negative. 
While a negative \apmf{} is unphysical, this insulates the analysis against the data preferring negative values, and thereby producing tighter limits. 
We find for \lcdm{}  + \apmf{}: $\sigma(\apmf) = XX$ (to be compared to YY from the Fisher matrix). 
We find for the 11-parameter model: $\sigma(\apmf) = XX$ (to be compared to YY from the Fisher matrix). 
This degree of consistency is reasonable. \tbd{}

>>>>

We present new limits on the amplitude of potential primordial magnetic fields (PMFs) using temperature and polarization measurements of the cosmic microwave background (CMB)  from \planck{}, \bicepkeck{}, \pb, and \sptpol. 
We reduce twofold the upper limit on the CMB anisotropy power due to a PMF, from $\apmf < 0.73$ from Planck alone to $\apmf < 0.35$ for the combined dataset at 95\% CL. 
%Most of the improvement is due to the addition of the \bicepkeck{} data; without these bandpowers, the combined limit weakens to $\apmf < 0.\tbd{XX}$. 
We also forecast the expected future limits from stage III CMB experiments (like \sptnew{},  \advactpol, or the \simons) and the proposed CMB stage IV experiment. 
Future CMB experiments should dramatically reduce the current uncertainties, by two orders of magnitude for stage III experiments and three orders of magnitude for stage IV. 
The constraints from a stage IV experiment have the potential to rule out much of the parameter space for PMFs.

@@

We find the addition of ground-based CMB polarization data to \planck{} data substantially improves the upper limits on the allowed power due to a PMF, leading to a factor of two improvement. 
With Planck alone, the 95\% confidence upper limit is $\apmf <  0.73$ for \lcdm{}+\apmf{}; with the ground-based experiments, this limit drops to $\apmf <  0.35$. 
The limits are slightly better with \alens{} allowed to vary (as a value for \alens{} slightly above unity is preferred). 
In this case for  Planck alone, the 95\% confidence upper limit is $\apmf <  0.68$, dropping to $\apmf <  0.27$ with the ground-based experiments. 
We test which datasets are important by adding single datasets to the \planck{} set while allowing \alens{} to float, and find that the majority of this improvement is due to the \bicepkeck{} ($\apmf <  0.28$) and \sptpol{} ($\apmf <  0.42$) B-mode power spectrum measurements. 
The other datasets do not have a large impact on the observed limits.

We propagate these limits on the observed PMF power into limits on the magnetic field strength, \bpmf. 
Due to the scaling with the fourth power, an upper limit with a flat prior on \apmf{} would lead to an apparent `detection' of \bpmf. 
We therefore importance sample the chains to generate a flat prior on \bpmf. 
We find a 95\% limit of $\bpmf < 1.97$\,nG for \planck{} alone, completely consistent with \citep{planck15-19} who found $\bpmf < 2.0$\,nG with $n_B=-2.9$ held fixed as we've done. 
This is slightly reduced to $\bpmf < 1.85$\,nG for the combined dataset. 
The improvement from adding more data is somewhat less than one might predict because the combined likelihood peaks about 1 sigma above zero, whereas the \planck{} only constraints peaks at zero.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:param_all} Parameter Constraints}
\small
\begin{tabular}{l | c c c }
Model   & $A_{pmf}$ &$A_{lens}$&$r$\\
\hline
\lcdm{} + \apmf{} &  $< 0.35$ & - & - \\
~~~+\alens{} & $< 0.27$ &$ 1.131 \pm  0.061$ &  -\\
~~~+\alens{}+r&  $< 0.24$&$ 1.133 \pm  0.062$ & $<0.056$\\ 
\end{tabular}
\tablecomments{ 
In the minimal model, \lcdm{} + \apmf{}, current CMB power spectra data from \planck, POLARBEAR, SPTpol, and \bicepkeck{} place a 95\% CL upper limit on the PMF power of $\apmf < 0.35$. 
This limit actually tightens slightly if tensor modes are allowed (r), or the amount of lensing is allowed to vary. 
In the case of r, the tightening is because the observed power at large angular scales is now split between PMFs or inflationary gravitational waves. 
In the case of \alens, it is because the preferred value for \alens{} is modestly above the predicted value of unity. 
Recall that the the PMF template is normalized to unity for \bpmf{} = 2.5\,nG.
} \normalsize
\end{center}
\end{table}

@@

\begin{table*}[tbh]
\begin{center}
\caption{\label{tab:experiments} Assumed survey parameters}
\small
\begin{tabular}{l || c c c c c }
Experiment & Sky coverage & Polarized Noise level  & 1/f knee & Beam FWHM \\
& &($\mu$K-arcmin)&&(arcmin.)\\
\hline
Stage III: & & & & \\

@@

%As a sanity check, we look at the Fisher matrix forecasts for Planck alone as well in both of these cases. 
%For both chains,  we look at the sigma in MCMC with \apmf{} allowed to go negative. 
%While a negative \apmf{} is unphysical, this insulates the analysis against the data preferring negative values, and thereby producing tighter limits. 
%We find for \lcdm{}  + \apmf{}: $\sigma(\apmf) = XX$ (to be compared to YY from the Fisher matrix). 
%We find for the 11-parameter model: $\sigma(\apmf) = XX$ (to be compared to YY from the Fisher matrix). 
%This degree of consistency is reasonable. \tbd{}
